NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2020-12-10 Manuscript/manuscriptPGPE.tex
Update manuscript

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameter magnitudes and signs of POPE and POPG
were determined by measuring the chemical-shift resolved dipolar splittings
with a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
%The corresponding order parameter signs were measured with a
S-DROSS experiments~\cite{gross97} using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
POPE and POPG powder were purchased from Avanti polar lipids.
The NMR experiments were identical to our previous work~\cite{antila19}.
\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
using the 2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.
For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13} and
the $\beta$-carbon peak was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm} (Fig. \ref{POPEspectra}).
For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
previously measured POPC spectra~\cite{ferreira13}, while $\alpha$ and  $\gamma$-carbon peaks
\todo{How were these assigned?} (Fig. \ref{POPGspectra}). The numerical value of the $\beta$-carbon
order parameter could not be determined, because its peak overlapped with the g$_2$ peak from glycerol backbone in POPG.
However, the order parameter of $\beta$-carbon is expected to be clearly smaller than for g$_2$
based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85}.
Therefore, the beginning of the S-DROSS curve gives the sign for g$_2$ order parameter and end for $\beta$ (Fig. \ref{POPGspectra} (E)).
This is confirmed with SIMPSON calculations using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
\todo{Details to be checked by Tiago}.

@@

Molecular dynamics simulation data were collected using
the Open Collaboration method \cite{botan15}, with
the NMR\-lipids Project blog (\url{nmrlipids.blogspot.fi}) and
GitHub repository (\url{github.com/NMRlipids/NMRlipidsIVotherHGs})
as the communication platforms.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

@@

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{../Figs/figure1.eps}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of {\bf A)} POPE  and {\bf B)} POPG.
    {\bf C)} Chemical structure of POPC and POPS.
    {\bf D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
   }
   \todo{This is a sketch, needs a lot of polishing.} \\
  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

@@

To experimentally characterize the differences in lipid headgroup conformational ensembles
in liquid bilayer phase, we measured the C-H bond order parameters magnitudes and signs
from POPG and POPE bilayers.
Order parameters for POPE are in good agreement with previous $^2$H NMR experiments \cite{seelig76}
and similar to our previous results for POPC \cite{ferreira16}.
In POPG, the $\beta$ and g$_2$ carbons have similar chemical environment and their peaks
overlap in the NMR spectra (Fig. \ref{HGorderParameters} B).
The signs of these order parameters were solved using SIMPSON simulations to interpre the
S-DROSS experiments in similar fashion as we did previously for PS lipids \cite{antila19}.
The resulting order parameters are compared with our previously published results
for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters} D.

%Our experimental order parameters, including the signs, from different lipid headgroups
%combined with the literature data are shown in Fig. \ref{HGorderParameters}.
%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.

The most distinct order parameters are observed for PS and PG headgroups.
In PS, the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
In PG headgroup,
%the $\alpha$-carbon order parameter is similar to PE and PG, while
the $\beta$-carbon order parameter has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.

@@

To characterize the differences in lipid headgroup conformational ensembles
leading to the distinct order parameters in PS and PG lipids, we calculate the
the distributions of heavy atom dihedral angles from CHARMM36 simulations.
Among available simulation models, this force field has least problems in
reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
The dihedral distributions in figure \ref{structures} show significant similarity between different lipids for all the other
bonds, except for the last two bonds in the headgroup end, O$_\alpha$-C$_\alpha$ and  C$_\alpha$-C$_\beta$.
These differences explaing the disctinct order parameters of
PS and PG headgroups with respenct to the other lipids.
The difference between PC and PE lipids in the O$_\alpha$-C$_\alpha$ bond dihedral
may be an artefact because the $\beta$-carbon order parameter in PC is poorly reproduced by the CHARMM36 force field \cite{botan15}.
The differences between lipids are reflected also to the 
angle between headgroup dipole and membrane normal,
which decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures}).

%
%As in previous NMRlipids project results for PC and PS lipids \cite{botan15,antila19},
%none of the MD simulation force fields correctly captures all the
%headgroup and glycerol backbone order parameters of PE and PG lipids
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG})
%that would enable a straightforward interpretation of conformational ensembles.
%
%show wide variation between different force fields
%and none of the force fields reproduce all values within experimental error bars.
%(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}),
%as observed previously also for PC and PS lipids \cite{botan15,antila19}.
%The Slipid simulations were able to capture the essential differences between PC and PS lipid
%headgroups \cite{antila19}, but this is not the case for PE and PG lipids.
%Without further discussion about poorly performing force fields, we focus on more detailed analysis
%of
%
%Nevertheless, CHARMM36 force field, which gives the results closest to the experiments for all lipids,
%captures the essential differences between
%PC, PS, PG and PE headroup order parameters  (Fig.~\ref{structures}) with the exception of $\beta$-carbon order parameter of PC
%
%give the best overall agreement with experiments for
%the headgroup and glycerol backbone order parameters of all PC, PS, PG and PE lipids in this and our previous
%studies \cite{botan15,antila19}. Even thought many order parameters in CHARMM36 simulations are
%not within the experimental error bars, it 
%
%which is too negative when compared with other lipids or experiments \cite{botan15}.
%
% As already pointed out in our previous study \cite{antila19},

The analysis of dihedral angle distributions reveal the free rotation of dihedral around
phosphorus-oxygen bonds in all lipids as all possible angles are observed in the distributions.
Results in figure \ref{structure} support the recently proposed models where this free rotation 
decouples the dynamics and conformational ensembles between acyl chains and headgroup in all lipids containing
the phosphorus group \cite{antila21}. 
However, some dihedral angles of phosphorus-oxygen bonds are less likely for PS, which
%suggests that the decoupling may be weaker for PS lipids
possibly explains the more rigid headgroup structures proposed for PS lipids~\cite{browning80,buldt81}.

@@

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
%the results suggest that free rotations around O-P bonds decouple the
%headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
%althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite the differences in dihedral distributions of two the last bonds at the headgroup end,
leading to the distinct order parameters for PS and PG lipids,
lipids with different headgroup sample dihedral angles within approximately same ranges.
Altogether, our results suggest that lipid headgroups and very flexible thereby being
able to adopt multiple conformations when interacting with proteins, ions or other biomolecules.
Furthermore, approximately same conformations are present liquid lamellar phase for different
headgroups studied here, althought the probability weight for the conformations are slightly different.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.

\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}
%Phosphatidylcholine headgroup conformational ensemble and orientation are unchanged
%upon addition of neutral molecules such as cholesterol or sphingomyelin into a bilayer,
%while
Order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup are known to depend
linearly on the incorporation of charged molecules into a lipid bilayer,
which is explained by tilting of the headgroup dipole orientation \cite{??}.
%are kn affect the headgroup dipole vector tilt which can
%be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons.
Such changes have been observed upon addition of charged lipids, proteins, surfactants,
drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.
%Headgroup order parameters of PC lipids decrease or increase upon addition of positive
%or negative charge into membrane, respectively, while they remain 
%\cite{??}.
%This has been explained by the tilting of the headgroup dipole due to electrostatic
%interactions with bound charges, and has been used to measure the binding affinity
%of ions to membranes \cite{??}.
%The headgroup order parameters of PC lipids can be used to measure the ion binding
%affinity to lipid bilayers because their magnitude is linearly proportional to the amount of bound charge in bilayer \cite{seelig87,catte16}.
%This molecular electrometer concept can be used also for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90,antila19}.
%The headgroup order parameters can be used to evaluate MD simulations against experimental data,
%because they can be directly calculated from MD simulations \cite{catte16}.

%\begin{figure}[]
%  \centering
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O3_M_M_G3C4_M_M_G3C5_M_M_G3N6_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3P2_M_M_G3O3_M_M_G3C4_M_M_G3C5_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3O1_M_M_G3P2_M_M_G3O3_M_M_G3C4_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G3_M_M_G3O1_M_M_G3P2_M_M_G3O3_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G2_M_M_G3_M_M_G3O1_M_M_G3P2_MCHARMM36.png}
%  \includegraphics[width=6.0cm]{../Figs/dihedralFIGS/POPCsurf_M_G1_M_M_G2_M_M_G3_M_M_G3O1_MCHARMM36.png}  
%  \caption{\label{DIHSwithSURF}
%
%  }
%\end{figure}

The experimentally measured decrease of PC headgroup order parameters upon addition of cationic surfactants into a bilayer
are well captured by CHARMM36 simulations in figure \ref{changesWITHsurf} A,
suggesting that these simulations can be used to interpret how lipid headgroup conformational ensembles
respond to the binding of positively charged into a membrane.
Charactization of conformational ensembles using heavy atom dihedral angle distributions in figure \ref{changesWITHsurf} B
reveal that choline region is essentially unchanged upon addition of carge.
The major changes upon addition of charge are observed in dihedrals related to the phosphate oxygens,
while small change is observed also in the g$_2$-g$_3$ bond in the glycerol backbone.
This result is in line with the recently proposed model suggesting that the flexible phosphate enables the headgroup
orient according to the charge accumulated into a membrane \cite{antila21}.

@@

Headgroup order parameters of PC lipids respond similarly also to the addition of charged lipids in experiments \cite{??}.
To resolve the conformational ensembles of lipid headgroups also in biologically relevant
mixtures, % of zwitterionic and charged lipids,
we compared the response of headgroup order parameters in
PC lipids to the addition of PE or PG (Fig. \ref{??}),
as well as in PG lipids to the addition of PC (Fig. \ref{??}),
between simulations and experiments. However, the accuracy of currently available force fields
is not sufficient to capture the headgroup conformational ensembles
in mixed membranes: 
The best performing force field for single componen membranes, CHARMM36, overestimates
the effect of PE to PC conformations and underestimates the response PC to the addition
of anionic PG headgroup. Previously we explained similar result for PC/PS mixtures  
%as we previously observed also for mixtures with PS lipids \cite{antila19}.
%The latter may arise from
by overbinding of the sodium counterions \cite{antila19}.

>>>>

\section{Methods}
\subsection{Experimental C--H bond order parameters}
The headgroup and glycerol backbone C--H bond order parameters of POPE and POPG
were measured using natural abundance $^{13}$C solid state NMR spectroscopy
as described previously \cite{ferreira13,ferreira16}.
The magnitudes of order parameters were determined from the chemical-shift resolved dipolar splittings
using a R-type Proton Detected Local Field (R-PDLF) experiment~\cite{dvinskikh04} and
the signs from S-DROSS experiments~\cite{gross97} combined with SIMPSON simulations \cite{bak00}.
%The experiments were done in a Bruker Avance III 400 spectrometer operating at a $^1$H Larmor frequency of 400.03 MHz.
%Magic angle spinning (MAS) of the sample was used at a frequency of 5.15 kHz (R-PDLF experiment) and 5 kHz (S-DROSS experiment).
%The following experimental setups were used.
The NMR experiments were identical as in our previous work~\cite{antila19}.
POPE and POPG powder were purchased from Avanti polar lipids \todo{Which isomer of POPG we have in the experiments?}.
%\todo{Is this enough and correct, or should we repeat some methods from the NMRlipidsIVps paper?}
The POPE experiments were recorded at 310~K and POPG experiments at 298~K, where the bilayers are in the liquid disordered phase \cite{marsh13}.

%Absolute values of the headgroup and glycerol backbone order parameters from PE and PG lipids are measured
%previously using $^2$H NMR~\cite{seelig76,gally81,wohlgemuth80,borle85}. Because also the order parameter
%signs bear essential information about the lipid structures \cite{botan15,ollila16}, we measured the
%magnitudes and signs of POPE and POPG C--H bond headgroup and glycerol backbone order parameter in liquid phase
%using the 2D-RPDLF and S-DROSS experiments, as described previously \cite{ferreira13,ferreira16,antila19}.

Glycerol backbone peaks from both lipids, and $\alpha$-carbon peak from POPE in the INEPT spectra 
were assigned based on previously measured POPC spectra~\cite{ferreira13}.
%For POPE, the glycerol backbone and $\alpha$-carbon peaks in INEPT spectra were assigned based on
%previously measured POPC spectra~\cite{ferreira13} and
The $\beta$-carbon peak from POPE was assigned based on $^{13}$C chemical shift table for amines available
at \url{https://www.chem.wisc.edu/areas/reich/nmr/c13-data/cdata.htm}.
%For POPG, the glycerol backbone peaks in INEPT spectra were assigned based on
%previously measured POPC spectra~\cite{ferreira13},
\todo{How were $\alpha$ and  $\gamma$-carbon peaks assigned in POPG?}.
The $\beta$-carbon peak from POPG overlapped with the g$_2$ peak from glycerol backbone
because their chemical environments are similar in POPG.
\todo{Details to be checked by Tiago}.

@@

Molecular dynamics simulation data were collected using
the NMRlipids Open Collaboration \cite{botan15}, which is running at
\url{nmrlipids.blogspot.fi} and \url{github.com/NMRlipids/NMRlipidsIVotherHGs}.
The simulated systems of pure PE and PG bilayers without additional ions
are listed in Tables \ref{systemsPE} and \ref{systemsPG},
and lipid mixtures with additional ions in Table~\ref{systemsMIX}.
Further simulation details are given in the SI, and
the simulation data are indexed in a
searchable database available at \url{www.nmrlipids.fi},
and in the NMRlipids/MATCH repository (\url{github.com/NMRlipids/MATCH}).

@@

%\clearpage
\begin{figure*}[]
  \centering
%  \includegraphics[width=9.0cm]{../Figs/lipids.pdf}
  %  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPCPSPEPG.eps}
   \includegraphics[width=18.0cm]{../Figs/figure1.eps}
   \caption{\label{HGorderParameters}
     Chemical structure, refocused-INEPT spectrum, 2D R-PDLF spectra, and S-DROSS data (from top to bottom) of {\bf A)} POPE  and {\bf B)} POPG.
      Full NMR spectra are shown Figs. \ref{POPEspectra} and \ref{POPGspectra}.
     {\bf C)} Chemical structure of POPC and POPS.
    {\bf D)} Headgroup and glycerol backbone order parameters from different experiments in lamellar liquid disordered phase.
    The values and signs for POPE (310~K) and POPG (298~K)
    measured in this work, and for POPS (298~K) \cite{antila19} and POPC (300~K) \cite{ferreira13,ferreira16}
    previously using $^{13}$C NMR. The literature values for
    DOPS with 0.1M of NaCl (303~K) \cite{browning80},
    POPG with 10nM PIPES (298~K) \cite{borle85},
    DPPG with 10mM PIPES and 100mM NaCl (314~K) \cite{wohlgemuth80}, 
    DPPE (341~K) \cite{seelig76},
    E.coliPE and E.coliPG (310~K) \cite{gally81}
    are measured using $^2$H NMR. The signs from $^{13}$C NMR are used also for the literature values.
   }
   \todo{This is a sketch, Tiago Ferreira will make a new figure.} \\
  \todo{D) could be clarified as Fig. 2 in the NMRlipids IVps paper.}
\end{figure*}

@@

To experimentally characterize the headgroup conformational ensembles of lipids
that are not bound to proteins in electrostatically neutral cell membrane,
we measured the C-H bond order parameters and their signs from of POPG and POPE
in liquid lamellar phase, as we did previously for POPC and POPS \cite{ferreira13,ferreira16,antila19}.
Determination of headgroup and glycerol backbone order parameters and their signs
was straightforward from the
%R-PDLF and S-DROSS
data in Figs. \ref{HGorderParameters}, \ref{POPEspectra} and \ref{POPGspectra}
for all the C-H bonds, except the $\beta$ and g$_2$ carbons in POPG.
These carbons have overlapping peaks in the INEPT spectra due to their similar chemical environments,
and only the magnitude of the larger order parameter could be determined from the R-PDLF spectra (Fig. \ref{HGorderParameters} B).
Nevertheless, based on previous $^2$H NMR measurements \cite{wohlgemuth80,gally81,borle85},
we assigned the larger order parameter to the g$_2$ carbon
%order parameter of $\beta$-carbon is expected to be clearly smaller than for 
and used the literature value for the $\beta$-carbon in SIMPSON simulations to determine the signs.
The decrease in the beginning of the S-DROSS curve suggests that the sign of larger g$_2$ order parameter
is negative and later increase suggests that sign of smaller $\beta$ order parameter is positive (Fig. \ref{HGorderParameters} (B)).
This interpretation is confirmed by SIMPSON calculations in Fig \ref{POPGsimpson}.
%using negative value for g$_2$ and positive value for $\beta$ order parameter (Fig. \ref{POPGsimpson}).
%The signs of these order parameters were solved using SIMPSON simulations to interpre the
%S-DROSS experiments in similar fashion as we did previously for PS lipids \cite{antila19}.

%Also the headgroup order parameters of PC lipids are close PE, althought the latter gives systematically
%slightly more positive values (Fig. \ref{HGorderParameters}).
% These could be explained with slightly larger temperature in PE measurements, except for the $\alpha$-carbon with the positive sign, for which the
% more positive value is farther away from zero.
%While the headgroup order parameters are similar for PC and PE lipids,
%PG and PS lipids exhibit distinct values in comparison with other lipids.

Experimental order parameters of POPC, POPE, POPG and POPS glycerol backbones and
headgroups from this and previous studies are collected in Fig. \ref{HGorderParameters} D),
where signs determined from $^{13}$C NMR experiments are used also for the $^2$H NMR data from the literature.
%The resulting order parameters are compared with our previously published results
%for POPC \cite{ferreira13,ferreira16} and POPS \cite{antila19} in figure \ref{HGorderParameters} D.
%Order parameters for POPE are in good agreement with previous $^2$H NMR experiments \cite{seelig76}
%and similar to our previous results for POPC \cite{ferreira16}.
Overall agreement of order parameters determined by different authors and different techniques
for same lipid headgroup is very good in here and previous studies \cite{botan15,ollila16,antila19},
suggesting that the observed differences between lipid types arise from differences in
headgroup chemistry rather than inaccuracies in experiments. 
The most distinct order parameters are observed for PS headgroups,
for which the $\alpha$-carbon order parameter exhibits significant forking
and the $\beta$-carbon has more negative value than in other lipids.
On the other hand, the $\beta$-carbon order parameter of PG headgroup
%the $\alpha$-carbon order parameter is similar to PE and PG, while
has positive sign in contrast to all the other lipids.
Notably, this has not been observed in traditional $^2$H NMR experiments,
%because absolute value of $\beta$-carbon order parameter is similar in PG, PE and PC lipids and
where only the absolute value of the order parameters are measured~\cite{wohlgemuth80,gally81,borle85}.
The glycerol backbone order parameters are similar for all the lipids, although they move slightly toward
positive values (closer to zero) in the order PC $<$ PE $<$ PS $<$ PG.
Essential differences between PC and PE headgroups are not observed.

@@

To resolve the conformational ensembles of lipid headgroups,
we compared the headgroup and glycerol backbone C-H bond order parameters
from different MD simulation force fields to the experimental data. 
None of the force fields reproduced the order parameters
within experimental accuracy for any of the lipids analysed in this work
(Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}.
%has least problems in
%reproducing the experimental lipid headgroup and glycerol backbone order parameters in different
%lipids \cite{botan15,antila19} (Figs. \ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}).
Nevertheless, the CHARMM36 force field performs best for all the headgroups and
%Importantly, it captures the experimentally observed distinct order parameters for PS and PG headgroups (Fig. \ref{??}).
%To characterize the differences in lipid headgroup conformational ensembles
reproduces the distinct order parameters of PS and PG lipids (Fig. \ref{structures} B).
To understand the structural origin of these distinct order parameters,
we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in figure \ref{structures} C).
%The dihedral distributions in figure \ref{structures} show significant similarity between different lipids for all the other
%bonds, except
Major differences between headgroups are observed only for the last two dihedrals
in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$,
which prefer {\it gauche$^-$} conformations for PG and {\it gauche$^+$} for PS.
Rest of the dihedrals are similar between different lipids, with the exception of
%However, some dihedral angles of phosphorus-oxygen bonds are less likely for PS, which
small differences for PS lipids. Therefore we suggest that the main differences between
lipid headgroups leading to distinct order parameter occur in the choline part, while
also changes in phosphate region may contribute in PS lipids, which could also
explain the more rigid headgroup structures~\cite{browning80,buldt81}.
%These differences explaing the disctinct order parameters of
%PS and PG headgroups with respenct to the other lipids.
%The differences between lipids are reflected also to
Furthermore, the angle between headgroup dipole and membrane normal
decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig. \ref{structures} B)).
However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P-N vector dipole may be an artefacts as the $\beta$-carbon order parameter in PC
is too negative the CHARMM36 force field, thereby not being equal to the order parameter
in PE as observed in experiments \cite{botan15}.

@@

In conclusion, all lipid headgroups sample very broad conformational ensembles in liquid lamellar phase.
%the results suggest that free rotations around O-P bonds decouple the
%headgroup and glycerol backbone structure and dynamics in similarly in all lipids,
%althought this rotation is slower in PS lipids where crossing certain barriers are slower.
Despite the differences in distributions
%of two the last bonds at the headgroup end,
%leading to the distinct order parameters for PS and PG lipids,
%lipids with
between headgroups, the sampled dihedral angles are within approximately same ranges for all headgroup types.
Therefore, we suggest that all lipid headgroups are very flexible thereby being
able to adopt wide range of multiple conformations when interacting with proteins, ions or other biomolecules.
%Furthermore, approximately same conformations are present liquid lamellar phase for different
%headgroups studied here, althought the probability weight for the conformations are slightly different.
%the order parameter experiments suggest that the glycerol backbone conformations in all lipids
%and the headgroup conformations in PC and PE lipids are relatively similar, while 
%PS and PG headgropus exhibit distinct conformational sampling.
%The details of sampled conformation are difficult to deduce from order parameters only,
%but the distinct headgroup order parameters
%of PS lipids are previously related to the more rigid structure of the headgroup~\cite{browning80,buldt81,antila19}.
%The analysis of dihedral angle distributions reveal the free rotation of dihedral around
%phosphorus-oxygen bonds in all lipids as all possible angles are observed in the distributions.
Our results support the models that propose free rotation around phosphate group for PC lipids,
which decouples the dynamics and conformatios between acyl chains and headgroups \cite{klauda08c,antila21b},
and suggests that these models could be applied for all lipids containing the phosphorus group.
The wide range of observed conformations suggest that the structures in lipid crystals \cite{buldt81,pascher92}
play only a minor role, and that the models aiming to explain NMR data using only few conformations \cite{seelig77c,davis83,Semchyschyn04,akutsu20}
are not sufficient to capture the large conformational space of lipids in liquid lamellar state.

\subsection{Lipid conformational ensembles in lipid bilayers with bound ions}

Charged lipids, proteins, surfactants, drugs, and ions incorporated in membranes
reorient the headgroup dipole in PC lipids, thereby affecting the order parameters of
lipid headgroups \cite{seelig87}. However, detailed understanding on
lipid conformational ensembles in membranes bearing electric charge is still lacking \cite{Semchyschyn04}.
%Order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup are known to depend
%linearly on the incorporation of charged molecules into a lipid bilayer,
%which is explained by tilting of the headgroup dipole orientation \cite{??}.
%%are kn affect the headgroup dipole vector tilt which can
%%be detected by measuring C-H bond order parameters of $\alpha$ and $\beta$ carbons.
%Such changes have been observed upon addition of charged lipids, proteins, surfactants,
%drugs, or ions \cite{??}, but how charges affect the lipid conformational ensemble remains unknown.

To resolve lipid headgroup conformational ensembles in cell membrane bearing positive charge,
%suggesting that these simulations can be used to interpret the charge induced changes in lipid conformational ensembles.
we first analyzed CHARMM36 simulations which correctly capture the experimentally measured decrease in PC headgroup order parameters
upon addition of cationic surfactants into a bilayer in figure \ref{changesWITHsurf} A).
Heavy atom dihedral angle distributions in figure \ref{changesWITHsurf} B
reveal that the decrease of trans state probabilities in g$_2$-g$_3$-O$_{g_3}$-P and g$_3$-O$_{g_3}$-P-O$_\alpha$
dihedrals are the major effects of positive charge on lipid headgroup conformations.
Choline region remains essentially unchanged and only minor changes are observed in other dihedrals
even thought almost half of the molecules in membranes are cationic surfactants.
%The major changes upon addition of charge are observed in dihedrals related to the phosphate oxygens,
%while small change is observed also in the g$_2$-g$_3$ bond in the glycerol backbone.
%This result is in line with the recently proposed model suggesting that the flexible phosphate enables the headgroup
%orient according to the charge accumulated into a membrane \cite{antila21}.
We also tried to resolve PC lipid headgroup conformational ensembles in mixed membrane with
negatively charged lipids \cite{scherer87},
%Headgroup order parameters of PC lipids respond similarly also to the addition of charged lipids in experiments \cite{??}.
%To resolve the conformational ensembles of lipid headgroups also in biologically relevant
%mixtures, % of zwitterionic and charged lipids,
%we compared the response of headgroup order parameters in
%PC lipids to the addition of PE or PG (Fig. \ref{??}),
%as well as in PG lipids to the addition of PC (Fig. \ref{??}),
%between simulations and experiments. However, the
but the accuracy of currently available force fields
was not sufficient for this (Fig. \ref{HGorderparametersPCvsPEPG}).
%to capture the headgroup conformational ensembles
%in mixed membranes: 
%The best performing force field for single componen membranes, CHARMM36, overestimates
%the effect of PE to PC conformations and underestimates the response PC to the addition
%of anionic PG headgroup. Previously we explained similar result for PC/PS mixtures  
%%as we previously observed also for mixtures with PS lipids \cite{antila19}.
%%The latter may arise from
%by overbinding of the sodium counterions \cite{antila19}.

@@

Also binding of ions may affect the zwitterionic and charged lipid headgroup conformational
ensembles in physiological conditions. The resolve this effect,
%In order to analyze how conformational ensemble of PC and PG lipids response to the
%bound ions,
we compared the changes of headgroup order parameters
in POPC:POPG mixtures 
upon addition of CaCl$_2$ between different simulations and experiments \cite{borle85,macdonald87}
in figures \ref{changesWITHCaClPG} (molar ratio 1:1) and \ref{changesWITHCaClPG1PC4} (molar ratio 4:1).
As in our previous studies \cite{catte16,melcr18,antila19,melcr20},  
%The decrease of POPC headgroup order parameters in mixtures with POPG lipids with
%increasing CaCl$_2$ concentration is overestimated in Slipids and Lipid17
%order parameters decrease in the response to bound cations in qualitative agreement with
%experiments, but
the calcium binding affinity to membranes is typically overestimated in simulations,
except by CHARMM36 with the NBfix correction which underestimates the binding affinity,
and the implicit inclusion of electronic polarizability improves the results.
Lipid17ecc model with the implicit inclusion of electronic polarizability gives
the most realistic response of PC lipid headgroup order parameters to the binding of calcium.
In this model, the main effect of calcium to the lipid conformational ensemble is
%the decrease is headgroup order parameters observed in this model is
the slight change of g$_3$-O$_{g3}$-P-O$_\alpha$ dihedral distribution from trans state
to eclipsed conformations (Fig. \ref{DIHSwithCAlipid17eccPOPC}).
This is in line with the changes observed in CHARMM36 simulations upon addition of charged surgactants (Fig. \ref{changesWITHsurf}),
despite the major differences in lipid headgroup conformational ensembles between
these models without ions (Fig. \ref{structures} vs. Fig. \ref{DIHSwithCAlipid17eccPOPC}).
%of lipid headgroup is not 
%Changes in lipid conformational ensembles characterized by distributions of heavy atom dihedral angles
%are very small upon binding of calcium or addition of cationic surfactants to membranes (Figs. \ref{??}).
%Yet, such changes are sufficient to tilt the headgroup dipole angle and reproduce the experimentally observed
%order parameter changes in $\alpha$ and $\beta$ carbons.
